---
title: "Data Wrangling Project"
author: "Tom Gocken"
date: "Tuesday, January 12, 2016"
output: html_document
---

### Step 1: Load datasets
- Data for counties of interest was queried from http://wonder.cdc.gov/EnvironmentalData.html
- Each environmental data type was downloaded in its own tab-delimited file and a dataset was created for each file.
```{r}
airtemp <- read.delim("data/Air Temperature.txt")
precip <- read.delim("data/Precipitation.txt")
sunlight <- read.delim("data/Sunlight.txt")
surfacetemp <- read.delim("data/Surface Temperature.txt")
```

### Step 2: Manage NA's
- **dplyr** package loaded for next steps:
```{r message=FALSE}
library(dplyr)
```

- Missing numeric values from certain columns in original files were populated with the string "Missing".
- "Missing" strings were converted to NA using **type.convert** function.

```{r}
airtemp <- mutate(airtemp, heat_index = 
                  type.convert(as.character(Avg.Daily.Max.Heat.Index..F.), 
                  na.strings = "Missing")
                  )                                          
surfacetemp <- mutate(surfacetemp, day_surface_temp = 
                  type.convert(as.character(Avg.Day.Land.Surface.Temperature..F.), 
                  na.strings = "Missing"),
                  night_surface_temp =
                  type.convert(as.character(Avg.Night.Land.Surface.Temperature..F.), 
                  na.strings = "Missing")
                  )
```

Note: The same result could have been accomplished using **gsub** function:

    airtemp <- mutate(airtemp, heat_index = as.numeric(gsub("Missing", NA, as.character((airtemp$Avg.Daily.Max.Heat.Index..F.)))))

### Step 3: Join data into a single tidy dataset
- Rows from individual datasets represent unique county/date observations. To ensure that all rows from every dataset were included in the combined dataset, full joins were used.
- Except for environmental data columns, variables were named consistently in downloaded files. As a result, natural joins on all variables with common names were possible without the need for a "by" argument.


```{r}
joindat <- full_join(airtemp, precip)
joindat <- full_join(joindat, sunlight)
joindat <- full_join(joindat, surfacetemp)
```
Date variable was created by concatenating year, month, and day columns and coverting to date class.
```{r}
joindat <- mutate(joindat, Date = as.Date(paste(joindat$Year.Code, joindat$Month.Code, joindat$Day.of.Month.Code, sep="-")))
```
Select statement used to assign consise variable names in common format to columms of interest.
```{r}
envdat <- select(joindat,
                 county = County,
                 year = Year,
                 day_of_yr = Day.of.Year,
                 date = Date,
                 max_air_temp = Avg.Daily.Max.Air.Temperature..F.,
                 min_air_temp = Avg.Daily.Min.Air.Temperature..F.,
                 heat_index,
                 precip = Avg.Daily.Precipitation..mm.,
                 sunlight = Avg.Daily.Sunlight..KJ.m².,
                 day_surface_temp,
                 night_surface_temp
                 )
```

### Step 4: Summarize and view data (UNDER CONSTRUCTION)
```{r}
summary(envdat)
```

...add some descriptive graphs...
